\Large \bf DRAFT: Direct-style process creation on Linux

### 2020-05-17 research/paper/paper.tex
paper

\begin{document}
\date{}
\title{\Large \bf DRAFT: Direct-style process creation on Linux}
% \author{
% {\rm Spencer Baugh}\\
% Two Sigma
% }
\maketitle
\begin{abstract}
Traditional process creation interfaces,
such as fork and spawn,
are complex to use, limited in power, and difficult to abstract over.
We develop a new process creation interface for Linux
which allows a program to create a child process in an non-running state
and initialize the new process by operating on it from the outside.
This method of process creation results in more comprehensible programs, 
has better error handling,
is more efficient,
and is more amenable to abstraction.
Our implementation is immediately deployable without kernel modifications on any recent Linux kernel version.
\end{abstract}

@@

We've implemented a variety of useful applications using direct-style process creation on Linux,
which we'll showcase in brief in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at Two Sigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable on recent Linux kernels.
\texttt{rsyscall} is open source and available from \url{https://github.com/catern/rsyscall}.

@@

We use this to create processes running the syscall request-response loop described in the previous section.
After this,
most system calls can be called as normal.
The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
Out of system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
Therefore we need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request will leave us waiting forever.

@@

In our fork-style implementations,
we assume a great deal of infrastructure is available to make fork-style simpler.
We assume the existence of a robust IPC system,
with an already-defined protocol which can cover all errors from all system calls.
Despite this, fork-style implementations are still significantly more verbose than direct-style.
\subsection{Clone versus fork}\label{microbench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{microbench}
 \caption{fork and clone performance under varying memory usage}
 \label{fig:microbench}
\end{figure}

We run a benchmark which times how long process creation takes
with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
with varying amounts of anonymous memory mapped in the parent process
before running the benchmark.
\texttt{clone} without \verb|CLONE_VM| functions identically to fork.
The results are in figure \ref{fig:microbench}.
Without \verb|CLONE_VM|,
process creation time is linear in the amount of memory mapped in the parent process.
At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
\texttt{fork} takes an average of 42.6 milliseconds to create a process,
while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.

This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
must set up memory mappings in the child process for all the memory mapped in the parent process.
\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
so its process creation time is constant in the amount of memory used in the parent process.

\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{subprocess_bench}
 \caption{Python subprocess and direct-style clone performance under varying memory usage}
 \label{fig:subprocess_bench}
\end{figure}

We benchmark Python's \texttt{subprocess.run} against direct-style \texttt{clone} with \texttt{rsyscall}
with varying memory usage, as in section \ref{microbench}.
The results are in Figure \ref{fig:subprocess_bench}.
TODO generate again to get the higher end where it crosses over.
\section{Related work}\label{related_work}
% TODO all this is just slapped onto the page without much thought, needs revising
\subsection{Direct-style process creation}
As we discuss in section \ref{background},
many academic operating systems have direct-style process creation.
We are not aware of any previous instances of direct-style process creation in a Unix-like environment.

@@

Some systems intercept system calls and forward them to another process to be implemented
as part of a virtualization system.
gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
In both systems,
system calls made in one process are intercepted,
and serviced by another process.
Also relevant is seccomp system call interception,
which can be used to intercept system calls and emulate them in another process.

Some systems forward system calls elsewhere for asynchronous execution.
FlexSC\cite{flexsc} is one example.
System calls were forwarded to dedicated system call execution threads.
The recently added Linux feature \verb|io_uring| is, in some sense, another example;
it supports sending a subset of system calls to the kernel,
which executes them asynchronously,
rather than blocking the thread.

Some systems call system calls in other processes for debugging or introspection purposes.
strace, gdb, and CRIU\cite{criu} are examples of this.
These systems typically uses ptrace,
which is capable of operating on multiple processes at once.
Unfortunately, as discussed in section \ref{implementation},
ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
a process which is already being ptraced cannot be ptraced again,
so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
\subsection{Process capabilities}
We mentioned the concept of process capabilities while discussing the design of \texttt{rsyscall} in section \ref{overview}.
The Capsicum system provides something called process descriptors\cite{capsicum},
which is a file descriptor handle for a process.
A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
These notions of process capability are quite limited, however;
pidfds and process descriptors only allow sending signals to a process or waiting for its death,
rather than exercising full control over the process.
\subsection{Capability-secure libc replacements}
The Capsicum system provides a set of system calls
which can be used to provide a capability-secure sandbox.\cite{capsicum}
Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
Programs ported to the capability-secure environment
are typically launched with an accompanying spawn-style wrapper.
This means that capability-secure processes are created primarily by an API
that is distinct from the capability-secure syscall API they have access to,
which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
Direct-style process creation could make it easier to create capability-secure processes 
from the capability-secure syscall API,
and reduce the need for a separate spawn-style wrapper.

Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
PLASH provides a spawn-style API (in the form of a shell)
to launch processes in a capability-secure environment.
PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
and is therefore limited in what kind of processes it can create.
\section{Future work}\label{future_work}
\subsection{Other applications of \texttt{rsyscall}}
Most avenues of future work focus on \texttt{rsyscall}.
\texttt{rsyscall} was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening \texttt{rsyscall}'s language support.
\subsection{Kernel support}
\texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

Several other aspects of our implementation would be improved by kernel support.
We discussed these in the implementation section;
in brief, we would most benefit from kernel support for
detecting when a child process finishes =execve=,
closing all =CLOEXEC= file descriptors except for an explicitly specified list,
and explicitly specifying a list of =CLOEXEC= file descriptors to inherit over =excve=.
Implementing these features in the kernel in a generally useful way, and upstreaming them,
is an important direction for our future work.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at Two Sigma.
While this gives us personally greater confidence in these techniques,
it would be better to use them in a publicly available, open source system.
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a meaningful demonstration of the viability of these techniques.
% \subsection{file descriptor lifetime management}
% % TODO this is hard to explain, probably best to drop this
% Keeping file descriptors open if and only if there is a specific process using that file descriptor
% is not the only possibility.
% Keep in mind that "a process using the file descriptor" is a slight abuse of terminology;
% processes don't use file descriptors, programs do, and in our system there is only one program.
% To reflect this, we could instead more directly use file descriptors in a process-agnostic way;
% this would support the creation of objects
% which work transparently across multiple processes which share a file descriptor table.
% Such objects would automatically use the relevant process to perform the syscalls for any specific operation.
% A process-centric view of file descriptors instead forces each object to be associated with one process.
% Nevertheless, we found that the process-centric perspective better matches the existing intuitions of users,
% especially those with prior experience in programming with processes.
% We hope that future systems for multi-process programming
% might explore an object-centric approach for managing resources.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.

>>>>

\begin{document}
\newcommand{\twosigma}{[tech company]} 
\date{}
\title{\Large \bf DRAFT: Direct-style process creation on Linux}
% \author{
% {\rm Spencer Baugh}\\
% Two Sigma
% }
\maketitle
\begin{abstract}
Traditional process creation interfaces,
such as fork and spawn,
are complex to use, limited in power, and difficult to abstract over.
We develop a new process creation interface for Linux
which allows a program to create a child process in an non-running state
and initialize the new process by operating on it from the outside.
This method of process creation results in more comprehensible programs, 
has better error handling,
is more efficient,
and is more amenable to abstraction.
Our implementation is immediately deployable without kernel modifications on any recent Linux kernel version.
\end{abstract}

@@

We've implemented a variety of useful applications using direct-style process creation on Linux,
which we'll demonstrate in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at \twosigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable on recent Linux kernels.
\texttt{rsyscall} is open source and available from
[Github url].
% \url{https://github.com/catern/rsyscall}.

@@

We use this to create processes running the syscall request-response loop described in the previous section.
After this,
most system calls can be called as normal.
The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
Out of system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}\label{execve}
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
Therefore we need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request will leave us waiting forever.

@@

In our fork-style implementations,
we assume a great deal of infrastructure is available to make fork-style simpler.
We assume the existence of a robust IPC system,
with an already-defined protocol which can cover all errors from all system calls.
Despite this, fork-style implementations are still significantly more verbose than direct-style.

We ensure that errors are reported back to the main program,
even without handling them;
this is a significant source of boilerplate in fork,
because reporting an error back to the main program requires IPC.
Even if there's no desire to handle errors,
proper error reporting is necessary for debugging.
And for serious users, proper error reporting and handling
is essential for a robust application.
% TODO hmm talk about... some stuff.
\subsection{Clone versus fork}\label{microbench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{microbench}
 \caption{fork and clone performance under varying memory usage}
 \label{fig:microbench}
\end{figure}

We run a benchmark which times how long process creation takes
with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
with varying amounts of anonymous memory mapped in the parent process
before running the benchmark.
Note that \texttt{clone} without \verb|CLONE_VM| is expected to function identically to fork.

All benchmarks are run pinned to an isolated core,
running on an Intel i9-9900K CPU with hyperthreading enabled.
Our benchmarks are intended to give a sense of relative, not absolute, performance,
and we have no reason to believe a more controlled benchmarking setup
would have differential effects on the performance of \texttt{fork} and \texttt{clone}.

The results are in figure \ref{fig:microbench}.
Without \verb|CLONE_VM|,
process creation time is linear in the amount of memory mapped in the parent process.
At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
\texttt{fork} takes an average of 42.6 milliseconds to create a process,
while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.

This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
must set up memory mappings in the child process for all the memory mapped in the parent process.
\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
so its process creation time is constant in the amount of memory used in the parent process.
\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
\begin{figure}[h!]
\centering
 \includegraphics[width=0.5\textwidth]{subprocess_bench}
 \caption{Python subprocess and direct-style clone performance under varying memory usage}
 \label{fig:subprocess_bench}
\end{figure}

We create a child process running \texttt{true} and wait for it to exit
using both the Python standard library's \texttt{subprocess.run},
and direct-style \texttt{clone} with \texttt{rsyscall}.
We benchmark the time each implementation takes
with varying amounts of mapped memory, as in section \ref{microbench}.
Both are running under CPython 3.7.7,
again pinned to an isolated core as with section \ref{microbench}.
The results are in Figure \ref{fig:subprocess_bench}.

The direct-style \texttt{clone} implementation stays around 2.2 milliseconds regardless of memory usage.
\texttt{subprocess.run}, which uses \texttt{fork},
starts at 1.4 milliseconds,
and then scales linearly with the amount of memory mapped, as expected.
Direct-style \texttt{clone} starts being faster than \texttt{subprocess.run}
at about 40 megabytes of memory mapped.

These results are acceptable.
Process creation in Python is already slow, taking milliseconds of time,
so it is not expected to be on the fast path.
In that context, \texttt{rsyscall} has a reasonable cost compared to \texttt{subprocess.run},
and avoids the bad scaling of \texttt{fork} which might be unexpected by the naive programmer.

Furthermore,
our implementation of \texttt{rsyscall} is barely-optimized pure-Python,
all the way down the syscall primitive,
whereas \texttt{subprocess.run} is entirely in C.
If we need to speed up our implementation,
we can perform some targeted optimizations,
reimplement some parts of the Python \texttt{rsyscall} library in native code,
or run our code using a faster Python interpreter such as PyPy.
\section{Related work}\label{related_work}
\subsection{Direct-style process creation}
As we discuss in section \ref{background},
many academic operating systems have direct-style process creation.
We are not aware of any previous instances of direct-style process creation in a Unix-like environment.

@@

Some systems intercept system calls and forward them to another process to be implemented
as part of a virtualization system.
gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
In both systems,
system calls made in one process are intercepted,
and serviced by another process.

Some systems forward system calls elsewhere for asynchronous execution.
FlexSC\cite{flexsc} is one example.
System calls in FlexSC are forwarded to a pool of dedicated system call execution threads.
The recently added Linux feature \verb|io_uring| is, in some sense, another example;
it supports sending a subset of system calls to the kernel,
which executes them asynchronously,
rather than blocking the thread.

Some systems call system calls in other processes for debugging or introspection purposes.
strace, gdb, and CRIU\cite{criu} are examples of this.
These systems typically uses ptrace,
which is capable of operating on multiple processes at once.
As discussed in section \ref{implementation},
a ptrace-based has approach has limitations which make it unsuitable for \texttt{rsyscall}.
\subsection{Process capabilities}
The Capsicum system provides something called process descriptors\cite{capsicum},
which is a file descriptor handle for a process.
A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
These notions of process capability are quite limited, however;
pidfds and process descriptors only allow sending signals to a process or waiting for its death,
rather than exercising full control over the process.
\subsection{Capability-secure libc replacements}
The Capsicum system provides a set of system calls
which can be used to provide a capability-secure sandbox.\cite{capsicum}
Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
developed this into a partial or full replacement for a POSIX libc.
Like all other Unix libcs that we know of,
these libc-replacements implicitly make system calls on the current process,
rather than using an explicit process capability as \texttt{rsyscall} does.

PLASH provides a spawn-style API, in the form of a shell,
to launch processes in a capability-secure environment.\cite{plash}
PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
and is therefore limited in what kind of processes it can create.
\section{Future work}\label{future_work}
\subsection{Other applications of \texttt{rsyscall}}
Most avenues of future work focus on \texttt{rsyscall}.
\texttt{rsyscall} was not developed solely for the purpose of this paper,
and it has other uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening \texttt{rsyscall}'s language support.
\subsection{Kernel support}
\texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct support in the Linux kernel
for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

Several other aspects of our implementation would also be improved by new features in the Linux kernel,
as discussed in sections \ref{execve} and \ref{cloexec}.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at \twosigma.
While this gives us personally greater confidence in these techniques,
it would be beneficial to use them in a publicly available, open source system.
Likely the best 
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a substantial demonstration of the viability of these techniques.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.
